---
title: "Analysis of Wildlife Strikes to Aircraft"
author: "Anusha Srinivasan"
date: "Spring 2024"
output:
  pdf_document: default
  html_document: default
subtitle: Practicum I CS5200
---
## Database Connection
This code block establishes connection to the database so that we can make queries directly from R and retrieve the data from the freemysqlhosting database.
```{r connectToMySQL, eval = TRUE, warning = FALSE, message = FALSE, echo = FALSE}
library(DBI)
library(RMySQL)

# Database connection details: Database Name, Host name, Password, Password, Port number
dbname <- "Database name"      
host <- "Host name"
user <- "User"       
password <- "Password"           

# Create connection to freemysqlhosting database.
con <- dbConnect(RMySQL::MySQL(),
                 dbname = dbname,
                 host = host,
                 username = user,
                 password = password,
                 port = port)
print("Connection successful")
#Disconnecting from the database
#DBI::dbDisconnect(con)
```

## Creating the Database Schema
This creates tables that will hold all the database schema. The database schema has four main tables:
- `flights table`: Records the flight details. It has unique fId, flight date, the origin airport, the airline name, the aircraft type, and a boolean indicating if it's a heavy aircraft type. It links to the `airports` table through the `originAirport` key.
- `airports table`: Records airport details. It has a unique aID, name, state, and a default Airport code.
- `conditions table`: Records sky conditions that can happen during flights and a text explanation for every sky condition.
- `strikes table`: Records bird strike incidents. It has a unique sid, and fid linking to flights table, number of birds with a default value, impact on the flight, damage, altitude in ft and it also links to the conditions table using cid.

```{r createDatabaseSchema, echo=FALSE, message=FALSE, warning=FALSE}
library(DBI)
#creating the flights table
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS flights (
    fId INT AUTO_INCREMENT PRIMARY KEY,
    date DATE,
    originAirport INT,
    airlineName VARCHAR(255),
    aircraftType VARCHAR(255),
    isHeavy BOOLEAN,
    FOREIGN KEY (originAirport) REFERENCES airports(aid)
  );
  ")
#creating the airports table
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS airports (
    aid INT AUTO_INCREMENT PRIMARY KEY,
    airportName VARCHAR(255),
    airportState VARCHAR(255),
    airportCode VARCHAR(3) DEFAULT 'ZZZ'
  );
")
#creating the conditions table
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS conditions (
    cid INT AUTO_INCREMENT PRIMARY KEY,
    sky_condition VARCHAR(255),
    explanation TEXT
  );
")
#creating the strikes table
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS strikes (
    sid INT AUTO_INCREMENT PRIMARY KEY,
    fid INT, 
    numbirds INT DEFAULT 1,
    impact TEXT,
    damage BOOLEAN,
    altitude INT CHECK (altitude >= 0),
    conditions INT,
    FOREIGN KEY (fid) REFERENCES flights(fId),
    FOREIGN KEY (conditions) REFERENCES conditions(cid)
  );
")
```
## Database Table Testing

This code block checks the existence of the tables: `flights`, `conditions`, `strikes`, and `airports`. If it exists, print its structure to see the columns in it.

## Checking the Flights Table
Check if the flights table exists and print the structure
```{r testFlightsTable, echo= FALSE,eval=FALSE}
#Check if the table exists
exists <- dbExistsTable(con, 'flights')
print(paste('Does flights table exist?', exists))
#If the table exists print the structure
if(exists) {
  structure <- dbGetQuery(con, 'DESCRIBE flights')
  print(structure)
}
```
## Checking the Conditions Table
Check if the Conditions table exists and print the structure
```{r testConditionsTable, echo= FALSE, eval=FALSE}
#Check if the table exists
exists <- dbExistsTable(con, 'conditions')
print(paste('Does conditions table exist?', exists))
#If the table exists print the structure
if(exists) {
  structure <- dbGetQuery(con, 'DESCRIBE conditions')
  print(structure)
}
```
## Checking the Strikes Table
Check if the Strikes table exists and print the structure
```{r testStrikesTable, echo= FALSE,eval=FALSE}
#Check if the table exists
exists <- dbExistsTable(con, 'strikes')
print(paste('Does strikes table exist?', exists))
#If the table exists print the structure
if(exists) {
  structure <- dbGetQuery(con, 'DESCRIBE strikes')
  print(structure)
}
```
## Checking the Airports Table
Check if the Airports table exists and print the structure
```{r testAirportsTable, echo=FALSE, eval=FALSE}
#Check if the table exists
exists <- dbExistsTable(con, 'airports')
print(paste('Does airports table exist?', exists))
#If the table exists print the structure
if(exists) {
  structure <- dbGetQuery(con, 'DESCRIBE airports')
  print(structure)
}
```
## Loading Bird Strike Data

This code block will load the bird strike data from a CSV file which has records of bird strike incidents like the date, location, airline, and impact of each incident. The CSV data is loaded into the dataframe `bds.raw`.

```{r loadCSV, echo=FALSE, message=FALSE, warning=FALSE}
# Load the csv birds data into a dataframe
bds.raw <- read.csv("BirdStrikesData-V3.csv")
print("CSV File loaded")
```
## Inserting data into airports table

This code block inserts value into the `airports` table with data from CSV file. Each airport has a unique aid, name and state, and a default code 'ZZZ' for the airport code.
```{r insertIntoAirportsTable, eval=TRUE, echo= FALSE,warning=FALSE}
insert_into_airports <- function(df, con) {
  #Get the unique airports
  unique_airports <- unique(df[, c("airport", "origin")])
  for (i in 1:nrow(unique_airports)) {
    #Get the airport name and state
    airport_name <- gsub("'", "''", unique_airports$airport[i], fixed = TRUE)
    airport_state <- gsub("'", "''", unique_airports$origin[i], fixed = TRUE)
    # Query to be executed where the airport code is default 'ZZZ'
    query <- sprintf("INSERT INTO airports (airportName, airportState, airportCode) VALUES ('%s', '%s', 'ZZZ') ON DUPLICATE KEY UPDATE airportName = airportName, airportState = airportState",
                     airport_name, airport_state)
    dbExecute(con, query)
  }
}
```
## Inserting data into conditions table

This code block inserts values into conditions table which has a list of sky conditions. It only adds unique values into the database.
```{r insertIntoConditionsTable, eval=TRUE,echo= FALSE, warning=FALSE}
insert_into_conditions <- function(df, con) {
  #Get the unique conditions
  unique_conditions <- unique(na.omit(df$sky_conditions))
  for (condition in unique_conditions) {
    if (condition != "") {
      # Query to be executed
      exist_query <- sprintf("SELECT COUNT(*) as Count FROM conditions WHERE sky_condition = '%s'",
                             gsub("'", "''", condition, fixed = TRUE))
      exist_result <- dbGetQuery(con, exist_query)
      
      if (exist_result$Count[1] == 0) {
        insert_query <- sprintf("INSERT INTO conditions (sky_condition) VALUES ('%s')",
                                gsub("'", "''", condition, fixed = TRUE))
        dbExecute(con, insert_query)
      }
    }
  }
}
```

## Preparing and Inserting Flight Data

The next two code blocks involve adding data from the CSV files and formatting date strings for SQL compatibility.

### Formatting Dates for SQL

This code block converts the date format directly compatible with SQL database. It takes a date string, attempts to parse it into a `Date` object and then format it into a string suitable for SQL.
```{r formatDateString, eval=TRUE, echo= FALSE,warning=FALSE}
format_date <- function(date_string) {
  if (is.na(date_string)) {
    return(NA)
  }
  tryCatch({
    # Change the format of date
    parsed_date <- as.Date(strptime(date_string, format = "%m/%d/%Y %H:%M"))
    if (!is.na(parsed_date)) {
      formatted_date <- format(parsed_date, "%Y-%m-%d")
    } else {
      return(NA)
    }
    return(formatted_date)
  }, error = function(e) {
    # Return NA if the date format is not correct
    return(NA)
  })
}
```

### Inserting Flight Records

The code block is used to insert the data into the flights table like the fId, date, origin airport, airline name, aircraft type, isHeavy Flag details.
```{r insertIntoFlightsTable, eval=TRUE,echo= FALSE, warning=FALSE}
insert_into_flights <- function(df, con) {
  # Skip if the airport name or state is unknown
  for (i in 1:nrow(df)) {
    if (df$origin[i] == "unknown" || df$airport[i] == "unknown") {
      next 
    }
    
    # Get the reference airport name and state
    airport_query <- sprintf("SELECT aid FROM airports WHERE airportName = '%s' AND airportState = '%s'",
                             gsub("'", "''", df$airport[i], fixed = TRUE),
                             gsub("'", "''", df$origin[i], fixed = TRUE))
    airport_result <- dbGetQuery(con, airport_query)
    
    if (nrow(airport_result) == 0) {
      warning(sprintf("No airport found for row %d.", i))
      next
    }
    
    airport_id <- airport_result$aid
    airline <- gsub("'", "''", df$airline[i], fixed = TRUE)
    aircraft <- gsub("'", "''", df$aircraft[i], fixed = TRUE)
    is_heavy <- ifelse(df$heavy_flag[i] == "Yes", 'TRUE', 'FALSE')

    # Change the format of the date
  flight_date <- format_date(df$flight_date[i])
    if (is.na(flight_date)) {
      warning(sprintf("Invalid date for row %d.", i))
      next
    }
  #Execute the query and insert flight_date, airport_id, airline, aircraft, is_heavy values
    query <- sprintf("INSERT INTO flights (date, originAirport, airlineName, aircraftType, isHeavy) VALUES ('%s', %d, '%s', '%s', %s)",
                     flight_date, airport_id, airline, aircraft, is_heavy)
    
    tryCatch({
      dbExecute(con, query)
    }, error = function(e) {
      warning(sprintf("Failed to insert for row %d: %s", i, e$message))
    })
  }
}
```

## Preparing and Inserting Strikes Data

The next three code blocks involve adding strikes data from the CSV files and retrieve the fId and cId from the flighst and conditions table.

### Lookup Flight ID Function
This code block queries the database to find the flight ID (fid) associated with a given airport name to linking strike incidents to flights based on the airport's name.
```{r lookupFid, eval=FALSE, echo= FALSE, warning=FALSE}
library(dplyr)
library(DBI)
lookup_fid <- function(con, airport_name) {
  #Query to be executed
  query <- sprintf("SELECT f.fid FROM flights f JOIN airports a ON f.originAirport = a.aid WHERE a.airportName = '%s' LIMIT 1",
                   gsub("'", "''", airport_name))
  result <- dbGetQuery(con, query)
  
  if (nrow(result) > 0) {
    # Return the first fid
    return(result$fid[1]) 
  } else {
    # Return NA if no fid found
    return(NA) 
  }
}
```

### Lookup Condition ID Function
This code block queries the database to find the condition ID (cid) based on sky conditions to link strike incidents with sky conditions recorded in the conditions table.
```{r lookupCid, eval=TRUE,echo= FALSE, warning=FALSE}
lookup_cid <- function(con, sky_conditions) {
  #Query to be executed
  query <- sprintf("SELECT cid FROM conditions WHERE sky_condition = '%s' LIMIT 1",
                   gsub("'", "''", sky_conditions)) # Escape single quotes for SQL
  result <- dbGetQuery(con, query)
  
  if (nrow(result) > 0) {
    # Return the first cid found
    return(result$cid[1]) 
  } else {
     # Return NA if no cid found
    return(NA)
  }
}
```

### Inserting Data into Strikes Table
This code block uses the lookup_fid and lookup_cid functions and inserts bird strike incident data into the strikes table. 
```{r insertIntoStrikesTable, eval=TRUE, echo= FALSE, warning=FALSE}

insert_into_strikes <- function(df, con) {
  for(i in 1:nrow(df)) {
     # Matches the name of the airport column
    fid <- lookup_fid(con, df$airport[i])
    # Matches the name of the sky conditions column
    cid <- lookup_cid(con, df$sky_conditions[i]) 

    # Skip the row if fid or cid is not found 
    if (is.na(fid) || is.na(cid)) {
      warning(sprintf("No matching flight ID or condition ID found for row %d", i))
      next 
    }
    impact <- gsub("'", "''", df$impact[i], fixed = TRUE)
    damage_value <- ifelse(tolower(df$damage[i]) == "damage", 'TRUE', 'FALSE')
    altitude <- ifelse(is.na(df$altitude_ft[i]) || df$altitude_ft[i] < 0, 0, df$altitude_ft[i])

    # Check for Altitude to be numeric
    if (!is.numeric(altitude)) {
      altitude <- as.numeric(altitude)
    }

    # Check altitude is not NA 
    if (is.na(altitude) || is.null(altitude)) {
      altitude <- 0 
    }
    numbirds <- 0
    # Ensure fid and cid are integers
    fid <- as.integer(fid)
    cid <- as.integer(cid)
    query <- sprintf("INSERT INTO strikes (fid, numbirds, impact, damage, altitude, conditions) VALUES (%s, %d, '%s', %s, %s, %s)",
                     ifelse(is.na(fid), "NULL", as.character(fid)), 
                     numbirds,  # Inserting default value for number of birds
                     impact, 
                     damage_value, 
                     ifelse(is.na(altitude), "NULL", as.character(altitude)), 
                     ifelse(is.na(cid), "NULL", as.character(cid)))
    tryCatch({
      dbExecute(con, query)
    }, error = function(e) {
      warning(sprintf("Failed to insert for row %d: %s", i, e$message))
    })
  }
}
```

## Function calls for inserting Data into the Tables

This code block has the function calls from the above code chunks to insert data into the CSV file
```{r insertDatafromCSV , eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# Replace empty strings or NA values with "unknown"
colnames(bds.raw)
bds.raw[bds.raw == ""] <- NA
bds.raw[is.na(bds.raw)] <- "unknown"
# Insert data into each table
insert_into_airports(bds.raw, con)
insert_into_conditions(bds.raw, con)
insert_into_flights(bds.raw, con)
insert_into_strikes(bds.raw, con)
```

## Testing the loading of the data into tables
This code block tests if the loading of data into the tables was successful by displaying the ifrst five rows of every table
```{r testloading, eval = FALSE , warning = FALSE, echo=FALSE}
# Display first 5 rows of all the tables
print(dbGetQuery(con, "SELECT * FROM airports LIMIT 5"))
 
print(dbGetQuery(con, "SELECT * FROM conditions LIMIT 5"))
 
print(dbGetQuery(con, "SELECT * FROM flights LIMIT 5"))
 
print(dbGetQuery(con, "SELECT * FROM strikes LIMIT 5"))
```
## Top Airports with Strike
This code block gives the top 10 states with the greatest number of bird strike incidents
```{r topAirportsWithStrike, eval = TRUE, warning = FALSE, echo=FALSE}
## Top 10 Airports with Strikes
query <- "
SELECT airportState, COUNT(*) AS numIncidents
FROM airports
JOIN flights ON airports.aid = flights.originAirport
JOIN strikes ON flights.fid = strikes.fid
GROUP BY airportState
ORDER BY numIncidents DESC
LIMIT 10
"
top_states <- dbGetQuery(con, query)
print(top_states)
```
## Analysis by Airline
This code block gives the top airlines that had an above average number bird strike incidents. 
```{sql connection=con, eval = T, warning = F, echo=FALSE}

SELECT airlineName, COUNT(*) AS NumberOfIncidents
  FROM flights
  JOIN strikes ON flights.fid = strikes.fid
  GROUP BY airlineName
  HAVING COUNT(*) > (
    SELECT AVG(IncidentCount) FROM (
      SELECT COUNT(*) AS IncidentCount
      FROM flights
      JOIN strikes ON flights.fid = strikes.fid
      GROUP BY airlineName
    ) AS AvgIncidents
  )
  ORDER BY NumberOfIncidents DESC;

```

## Analysis by month
This code block gives the total number of wildlife strikes by month.
```{r analysisByMonth, eval = TRUE, warning = FALSE, echo=FALSE}
library(dplyr, warn.conflicts = FALSE)
query <- "
SELECT MONTH(date) AS month, COUNT(*) AS num_strikes
FROM flights
GROUP BY MONTH(date)
ORDER BY MONTH(date)
"
#Dataframe is used to store the query
monthly_strikes <- dbGetQuery(con, query)
# Converting the month numbers to month names
monthly_strikes$month <- month.abb[monthly_strikes$month]
print(monthly_strikes)
```
## Trend by month
This code block gives how wildlife strikes vary by month
```{r trendByMonth,fig.align='center' eval = TRUE, warning = FALSE, echo=FALSE}
months_ordered <- factor(monthly_strikes$month, levels = month.abb)
# Generate bar plot
bp <- barplot(monthly_strikes$num_strikes, names.arg = months_ordered, 
              main = "Strikes by Month", 
              xlab = "Month", 
              ylab = "Number of Strikes",
              col = "green")
text(x = bp, y = monthly_strikes$num_strikes, labels = monthly_strikes$num_strikes, pos = 3, cex = 0.8)


```

## Audit Logging 
This code block creates an audit_log table designed to track modifications within the database, specifically deletions from the strikes table.
```{r storedProcedure, echo=FALSE}
# Creating a Audit Log Table
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS audit_log (
    id INT AUTO_INCREMENT PRIMARY KEY,
    modification_type VARCHAR(255),
    modified_table VARCHAR(255),
    modification_timestamp DATETIME,
    record_id INT,
    description TEXT
  )
")
dbExecute(con, "DROP PROCEDURE IF EXISTS RemoveStrike")
# Creating a Stored Procedure for Removing a Strike
dbExecute(con, "
  CREATE PROCEDURE RemoveStrike(IN strikeID INT)
  BEGIN
    INSERT INTO audit_log (modification_type, modified_table, modification_timestamp, record_id, description)
    VALUES ('Deletion', 'strikes', NOW(), strikeID, CONCAT('Deleted strike with ID ', strikeID));

    DELETE FROM strikes WHERE sid = strikeID;
  END
")

# Invoke the stored procedure with a specific strike ID to remove
strikeIDToRemove <- 1 
dbExecute(con, sprintf("CALL RemoveStrike(%d)", strikeIDToRemove))

# Get recent entries in the audit log
auditLog <- dbGetQuery(con, "SELECT * FROM audit_log ORDER BY modification_timestamp DESC LIMIT 5")
print(auditLog)

#Test to check if the function
latest_audit_log_entry <- dbGetQuery(con, "
SELECT * FROM audit_log 
WHERE record_id = @test_strike_id 
ORDER BY modification_timestamp DESC LIMIT 1;
")
print(latest_audit_log_entry)
```

